---
output: html_document
title: "The many roads to a statistical gene-environment interaction"
bibliography: ../doc/library.bib
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F, message = F,warning = F, 
                      fig.path = "../output/gxe_types_viz/", dpi = 300)
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "patchwork", "cowplot"), 
  library, character.only = TRUE))
theme_set(theme_bw())
theme_update(axis.ticks = element_blank(), axis.text = element_blank(),
             legend.title.align=0.5)
scale_color_discrete <- function(...) scale_color_brewer(..., palette = "Set1")
scale_fill_discrete <- function(...) scale_fill_brewer(..., palette = "Set1")
```

# Phenomena leading to statistical interactions

```{r phenomena}
full_names <- c(
  func = "Functional", 
  nlm = "Nonlinear mediator", 
  rge_nl = "G-E correlation plus nonlinearity",
  hv = "Heterogeneous variability / vQTL", 
  hm = "Heterogeneous measurement"
)
```

```{r viz-prep}
present_lm_fit <- function(df, form_str = "y ~ g * e") {
  lm(as.formula(form_str), data = df) %>%
    broom::tidy() %>%
    filter(term != "(Intercept)") %>%
    select(term, estimate, p.value) %>%
    mutate(across(c(estimate, p.value), ~round(., 3)))
}

set.seed(123)

n_sim <- 10000
base_sim_df <- tibble(.rows = n_sim)
base_sim_df$g <- rbinom(n_sim, 2, 0.25)
base_sim_df$g_fac <- factor(base_sim_df$g, labels = c("AA", "AB", "BB"))

toy_diagrams <- list()
sim_plots <- list()
lm_fits <- list()
```

## Functional

```{r functional, fig.asp=0.5}
func_sim_df <- base_sim_df %>%
  mutate(
    e = rnorm(n_sim, 5, 1),
    y = rnorm(n_sim, e + g * e, 1)
  )

func_cartoon <- ggdraw() + 
  draw_image(magick::image_read_pdf("../output/functional_cartoon.pdf", density = 600))

toy_diagrams$func <- func_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y") +
  guides(color = "none")

sim_plots$func <- toy_diagrams$func +
  geom_point(alpha = 0.2) +
  guides(color = "legend")

lm_fits$func <- present_lm_fit(func_sim_df)

func_cartoon + toy_diagrams$func + sim_plots$func +
  plot_layout(widths = c(2, 1, 1)) +
  plot_annotation(tag_levels = "a")
```

## Nonlinear mediator of genes and evironment

```{r nonlinear-mediator, fig.asp=0.4}
cube_root <- function(x) {
  sign(x) * abs(x) ^ (1/3)
}
  
nlm_sim_df <- base_sim_df %>%
  mutate(
    e = rnorm(n_sim, 0, 1),
    m = rnorm(n_sim, g + e, 0.1),
    y = rnorm(n_sim, cube_root(m + 4), 0.1)
  )

toy_diagrams$nlm1 <- nlm_sim_df %>%
  ggplot(aes(x = e, y = m, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  labs(x = "E", y = "M") +
  guides(color = "none")
toy_diagrams$nlm2 <- nlm_sim_df %>%
  ggplot(aes(x = m, y = y)) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 2)", 
              color = "black", se = FALSE) +
  labs(x = "M", y = "Y")

sim_plots$nlm <- nlm_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y")

lm_fits$nlm <- present_lm_fit(nlm_sim_df)

toy_diagrams$nlm1 + toy_diagrams$nlm2 + sim_plots$nlm +
  plot_layout(widths = c(1, 1, 2)) +
  plot_annotation(tag_levels = "a")
```

## G-E correlation plus nonlinearity

```{r ge-nonlinear, fig.asp=0.5}
rge_nl_sim_df <- base_sim_df %>%
  mutate(
    e = rnorm(n_sim, g, 1),
    y = rnorm(n_sim, cube_root(e + 4), 0.1)
  )

seg_df <- tibble(
  x = c(-2, 0, 2),
  y = cube_root(x + 4),
  slope = 1/3 * (x + 4) ^ (-2/3),
  intercept = y - x * slope,
  x_start = x - 2,
  x_end = x + 2,
  y_start = slope * x_start + intercept,
  y_end = slope * x_end + intercept
)

toy_diagrams$rge_nl1 <- rge_nl_sim_df %>%
  ggplot(aes(x = g_fac, y = e, color = g_fac)) +
  geom_boxplot() +
  guides(color = "none") +
  scale_color_discrete() +
  labs(x = "G", y = "E")

toy_diagrams$rge_nl2 <- rge_nl_sim_df %>%
  ggplot(aes(x = e, y = y)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), 
              se = FALSE, color = "black", fullrange = TRUE) +
  geom_segment(aes(x = seg_df$x_start[1], xend = seg_df$x_end[1],
                   y = seg_df$y_start[1], yend = seg_df$y_end[1]),
               linetype = "dashed", color = RColorBrewer::brewer.pal(3, "Set1")[1], size = 1.25) +
  geom_segment(aes(x = seg_df$x_start[2], xend = seg_df$x_end[2],
                   y = seg_df$y_start[2], yend = seg_df$y_end[2]),
               linetype = "dashed", color = RColorBrewer::brewer.pal(3, "Set1")[2], size = 1.25) +
  geom_segment(aes(x = seg_df$x_start[3], xend = seg_df$x_end[3],
                   y = seg_df$y_start[3], yend = seg_df$y_end[3]),
               linetype = "dashed", color = RColorBrewer::brewer.pal(3, "Set1")[3], size = 1.25) +
  labs(x = "E", y = "Y")

sim_plots$rge_nl <- rge_nl_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y")

lm_fits$rge_nl <- present_lm_fit(rge_nl_sim_df)

toy_diagrams$rge_nl1 + toy_diagrams$rge_nl2 + sim_plots$rge_nl +
  plot_layout(widths = c(2, 3, 3)) +
  plot_annotation(tag_levels = "a")
```

## Heterogeneous variability

```{r hv, fig.asp=0.5}
hv_sim_df <- base_sim_df %>%
  mutate(
    e = rnorm(n_sim, 5, 1),
    y_star = rnorm(n_sim, scale(e), 1),  # Incorporates both noise and the E-informed percentile within that noise
    y = rnorm(n_sim, g * y_star, 1)  # Noise has already been incorporated into y_star
  )

toy_diagrams$hv1 <- hv_sim_df %>%
  ggplot(aes(x = g_fac, y = y, color = g_fac, group = g_fac)) +
  geom_boxplot() +
  geom_segment(aes(x = 0.75, xend = 1.25, y = -0.4, yend = -0.4, linetype = "dashed")) +
  geom_segment(aes(x = 0.75, xend = 1.25, y = 0.4, yend = 0.4, linetype = "dotted")) +
  geom_segment(aes(x = 1.75, xend = 2.25, y = -0.8, yend = -0.8, linetype = "dashed")) +
  geom_segment(aes(x = 1.75, xend = 2.25, y = 0.8, yend = 0.8, linetype = "dotted")) +
  geom_segment(aes(x = 2.75, xend = 3.25, y = -1.5, yend = -1.5, linetype = "dashed")) +
  geom_segment(aes(x = 2.75, xend = 3.25, y = 1.5, yend = 1.5, linetype = "dotted")) +
  scale_linetype_identity(breaks = c("dashed", "dotted"), labels = c("E1", "E2"), 
                          name = "E", guide = "legend") +
  scale_color_discrete() +
  guides(color = "none") +
  labs(x = "G", y = "Y")
toy_diagrams$hv2 <- hv_sim_df %>%
  ggplot(aes(x = e, y = y_star, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  # scale_x_continuous(breaks = c(-0.5, 0.5), labels = c("E1", "E2")) +
  geom_vline(xintercept = 3, linetype = "dotted") +
  geom_vline(xintercept = 7, linetype = "dashed") +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y*") +
  # theme(axis.ticks.x = element_line(color = "red", size = 2)) +
  guides(color = "none")
toy_diagrams$hv3 <- hv_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  geom_vline(xintercept = 3, linetype = "dotted") +
  geom_vline(xintercept = 7, linetype = "dashed") +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y") +
  guides(color = "none")

sim_plots$hv <- hv_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y")

lm_fits$hv <- present_lm_fit(hv_sim_df)

toy_diagrams$hv1 + (toy_diagrams$hv2 / toy_diagrams$hv3) + sim_plots$hv +
  plot_layout(widths = c(3, 2, 3)) +
  plot_annotation(tag_levels = "a")
```

## Heterogeneous measurement

```{r heterogeneous-measurement, fig.asp=0.5}
hm_sim_df <- base_sim_df %>%
  mutate(
    e = rnorm(n_sim, 5, 1),
    e_m = rnorm(n_sim, e, exp(0.5 * g)),
    y = rnorm(n_sim, e, 1)
  )

icc_df <- hm_sim_df %>%
  group_by(g_fac) %>%
  summarise(icc = var(e) / var(e_m))

toy_diagrams$hm1 <- icc_df %>%
  ggplot(aes(x = g_fac, y = icc, fill = g_fac)) +
  geom_bar(stat = "identity") +
  guides(fill = "none") + 
  labs(x = "G", y = "Exposure ICC (measurement accuracy)") +
  theme(axis.ticks.y = element_line())
toy_diagrams$hm2 <- hm_sim_df %>%
  ggplot(aes(x = e, y = y, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  scale_color_discrete(name = "G") +
  labs(x = "E (true)", y = "Y") +
  guides(color = "none")
toy_diagrams$hm3 <- hm_sim_df %>%
  ggplot(aes(x = e_m, y = y, color = g_fac, group = g_fac)) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, fullrange = TRUE) +
  scale_color_discrete(name = "G") +
  labs(x = "E (measured)", y = "Y") +
  guides(color = "none")

sim_plots$hm <- hm_sim_df %>%
  ggplot(aes(x = e_m, y = y, color = g_fac, group = g_fac)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
  scale_color_discrete(name = "G") +
  labs(x = "E", y = "Y")

lm_fits$hm <- present_lm_fit(hm_sim_df, "y ~ g * e_m")

toy_diagrams$hm1 + (toy_diagrams$hm2 / toy_diagrams$hm3) + sim_plots$hm +
  plot_layout(widths = c(2, 2, 3)) +
  plot_annotation(tag_levels = "a")
```

```{r heterogeneous-covariate-measurement}
hcm_sim_df <- base_sim_df %>%
  mutate(
    z = rnorm(n_sim, 5, 1),
    e = rnorm(n_sim, z, 1),
    z_m = rnorm(n_sim, z, 0.5 * exp(2 - g)),
    y = rnorm(n_sim, e + z, 1)
  )
present_lm_fit(hcm_sim_df %>% filter(g == 0), "y ~ e + z_m")
present_lm_fit(hcm_sim_df %>% filter(g == 1), "y ~ e + z_m")
present_lm_fit(hcm_sim_df %>% filter(g == 2), "y ~ e + z_m")
present_lm_fit(hcm_sim_df, "y ~ g * e + g*z_m")

lm_fits$hcm <- present_lm_fit(hcm_sim_df, "y ~ g * z_m")
```

# Appendix: Simulation setup

1. Simulate G with MAF = 0.25: $G \sim binom(2, 0.25)$ ($N_{sims}$ = 10,000)
2. Simulate E, M (where applicable), and Y as follows:

| GxE Type | Exposure (E) | Mediator (M) | Outcome (Y) |
| :---: | :---: | :---: | :---: |
| Functional | $N(5, 1)$ |  | $N(E + G*E, 1)$ |
| Nonlinear mediator | $N(3, 1)$ | $N(G + E, 0.25)$ | $N(\sqrt{M}, 0.1)$ |
| vQTL | $N(5, 1)$ | | $Y^* \sim N(std(E), 1)$ <br /> $Y \sim N(G * Y^*, 1)$ |
| Heterogeneous measurement | $N(5, 1)$ <br /> $E_m \sim N(E, 0.5 * e^{2-G})$ | | $Y \sim N(E, 1)$ <br /> $Y_m \sim N(Y, 0.5 * e^G)$ |
| G-E correlation with nonlinearity | $N(3-G, 1)$ | | $N(\sqrt{E}, 1)$ |

3. Test for statistical interaction using the following regression model:
$$
Y \sim G + E + G*E
$$

```{r simulation-results}
lm_fits %>%
  bind_rows(.id = "phenomenon") %>%
  filter(phenomenon != "hcm") %>%
  mutate(term = toupper(term)) %>%
  select(Phenomenon = phenomenon, `Regression term` = term, 
         `Effect estimate` = estimate, `P-value` = p.value) %>%
  write_csv("../output/simulation_results.csv")

lm_fits %>%
  bind_rows(.id = "phenomenon") %>%
  filter(phenomenon != "hcm") %>%
  kbl(caption = "Regression results from simulations") %>%
  kable_paper() %>%
  column_spec(2, width = "1cm") %>%
  pack_rows(full_names[1], 1, 3) %>%
  pack_rows(full_names[2], 4, 6) %>%
  pack_rows(full_names[3], 7, 9) %>%
  pack_rows(full_names[4], 10, 12) %>%
  pack_rows(full_names[5], 13, 15)
```
